---
title: "Human Activity Recognition Model"
author: "Icaro Oliveira de Oliveira"
date: "Saturday, August 22, 2014"
output:
  html_document
---
# Contextualization
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

The aim of this report was to use data from accelerometers placed on the belt, forearm, arm, and dumbell of six participants to predict how well they were doing the exercise in terms of the classification in the data. 

### Libraries
The following libraries were used throughout the code.
```{r}
library(knitr)
library(ROCR)
library(caret)
library(randomForest)
```
```{r setoptions, echo = FALSE}
opts_chunk$set(cache = FALSE)
```

### Loading and preprocessing the data
Two csv files contatining the training and test data was downloaded from Amazon's cloudfront on the 22/08/2014 into a data folder in the working directory. 

```{r, eval = FALSE}
# check if a data folder exists; if not then create one
if (!file.exists("data")) {dir.create("data")}

# file URL and destination file
fileUrl1 <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
destfile1 <- "./data/pml-training.csv"
fileUrl2 <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
destfile2 <- "./data/pml-testing.csv"

# download the file and note the time
if (!file.exists(destfile1)) {download.file(fileUrl1, destfile = destfile1)}
if (!file.exists(destfile2)) {download.file(fileUrl2, destfile = destfile2)}
```

The training data was then loaded into R.

```{r}
# read the csv file for training 
data <- read.csv("./data/pml-training.csv", na.strings= c("NA",""," "))
```

There was a lot of NAs values in the data which would create a lot of noise for the model. As a result, these columns were removed from the data set. The first eight columns that represented identifiers for the experiment were also removed.

```{r}
data_NAs <- apply(data, 2, function(x) {sum(is.na(x))})
data_clean <- data[,which(data_NAs == 0)]
data_clean <- data_clean[8:length(data_clean)]
```

### Creating a model
The test data set was split up into 70% to training and 30% to cross validation sets.

```{r}
inTrain <- createDataPartition(y = data_clean$classe, p = 0.7, list = FALSE)
training <- data_clean[inTrain, ]
crossval <- data_clean[-inTrain, ]
```
A model was fitted with the outcome set to the training class and all the other variables used to predict.

```{r}
model <- randomForest(classe ~ ., data = training)
model
```

The model produced a very small OOB error rate. This was deemed satisfactory enough to progress the testing.

### Cross-validation
The model was then used to classify the remaining 30% of data. The results were placed in a confusion matrix along with the actual classifications in order to determine the accuracy of the model.

```{r}
predictCrossVal <- predict(model, crossval)
confusionMatrix(predictCrossVal,crossval$classe)
```

This model yielded a 99.5% prediction accuracy. Again, this model proved very robust and adequate to predict new data.

### Predictions
A separate data set was then loaded into R and cleaned in the same manner as before. The model was then used to predict the classifications of the 20 different test cases.

```{r}
# apply the same treatment to the final testing data
data_test <- read.csv("./data/pml-testing.csv", na.strings= c("NA",""," "))
data_test_NAs <- apply(data_test, 2, function(x) {sum(is.na(x))})
data_test_clean <- data_test[,which(data_test_NAs == 0)]
data_test_clean <- data_test_clean[8:length(data_test_clean)]

# predict the classes of the test set
predictTest <- predict(model, data_test_clean)
predictTest
```

### Conclusions
With the abundance of information given from multiple measuring instruments, it's possible to accurately predict how well a person is performing an exercise using a simple model.